---
title: "Mini-Project 02 - How Do You Do 'You Do You'?"
author: "miggonzalz"
editor:
    mode: source
format:
  html:
    code-fold: true
    toc: true            
    number-sections: true #
---

## Introduction
In this mini project, I analyze microdata from the [**American Time use Survey (ATUS)**](https://www.bls.gov/tus/) to explore how specific subgroups of Americans spend their daily time. With the help of R, I used **dplyr** to **clean, merge, and summarize data** from multiple data sources, then used **ggplot2** to **visualize key trends** seen across demographics. 

For the **final deliverable**, I am acting as product manager for a start-up launching a **time-tracking app**. Unlike typical calendar apps, this will compare users against demographically similar peers (e.g,. "Congrats! You exercised 3 more hours this week than others like you!") Using gamification to encourage healthier time-use habits. 


## Data Acquisition and Preparation

The following code outlines the steps I have taken to upload the data from ATUS alongside some data manipulation steps to complete this mini project.


```{r}
library(tidyverse)
library(glue)
library(readxl)
library(httr2)
library(lubridate)
library(gt)
library(scales)

# --- FUNCTION: Load and Clean ATUS Data ---
load_atus_data <- function(file_base = c("resp", "rost", "act")){
    file_base <- match.arg(file_base)
    file_name_out <- file.path("data", "mp02", glue("atus{file_base}_0324.dat"))
    
    if(!file.exists(file_name_out)){
        stop(glue("File {file_name_out} not found. Please ensure data is downloaded."))
    }
    
    # Read data forcing IDs and Codes to Character
    file <- read_csv(file_name_out, 
                     show_col_types = FALSE,
                     col_types = cols(TUCASEID = col_character(),
                                      TRCODEP = col_character()))
    
    result <- switch(file_base, 
        resp = file |> 
            mutate(participant_id = TUCASEID,
                   survey_weight = TUFNWGTP,
                   survey_year = TUYEAR,
                   diary_date = ymd(TUDIARYDATE),
                   diary_month = month(diary_date, label = TRUE),
                   time_alone = TRTALONE,
                   childcare_min = TRTCC + TRTCCTOT,
                   family_min = TRTHHFAMILY,
                   spouse_min = TRTSPOUSE,
                   children_count = TRCHILDNUM,
                   employment_status = case_match(TELFS,
                       1:2 ~ "Employed",
                       3:4 ~ "Unemployed", 
                       5 ~ "Not in Labor Force",
                       .default = "Unknown"
                   ),
                   hh_child_under_18 = case_match(TRHHCHILD, 1 ~ 1, 2 ~ 0, .default = 0),
                   weekly_hours = TESPUHRS,
                   hourly_wage = TEERNH1O,
                   industry_code = TRDTIND1,
                   industry_desc = case_match(TRDTIND1,
                       1 ~ "Agriculture", 
                       2 ~ "Forestry/Fishing", 
                       3 ~ "Mining", 4 ~ "Construction",
                       5 ~ "Nonmetallic manufacturing", 
                       6 ~ "Primary metals manufacturing",
                       7 ~ "Machinery manufacturing",
                       8 ~ "Computer/electronic manufacturing",
                       9 ~ "Electrical manufacturing", 
                       10 ~ "Transportation manufacturing",
                       11 ~ "Wood products",
                       12 ~ "Furniture manufacturing",
                       13 ~ "Miscellaneous manufacturing",
                       14 ~ "Food manufacturing",
                       15 ~ "Beverage/tobacco manufacturing", 
                       16 ~ "Textile/apparel manufacturing",
                       17 ~ "Paper/printing", 
                       18 ~ "Petroleum/coal manufacturing",
                       19 ~ "Chemical manufacturing", 
                       20 ~ "Plastics/rubber manufacturing",
                       21 ~ "Wholesale trade", 
                       22 ~ "Retail trade",
                       23 ~ "Transportation/warehousing", 
                       24 ~ "Utilities",
                       25 ~ "Publishing", 
                       26 ~ "Motion picture/sound recording",
                       27 ~ "Broadcasting", 28 ~ "Internet publishing",
                       29 ~ "Telecommunications", 30 ~ "Internet/data processing",
                       31 ~ "Other information services", 32 ~ "Finance",
                       33 ~ "Insurance", 34 ~ "Real estate", 35 ~ "Rental/leasing",
                       36 ~ "Professional/technical services", 37 ~ "Management companies",
                       38 ~ "Administrative services", 39 ~ "Waste management",
                       40 ~ "Educational services", 41 ~ "Hospitals",
                       42 ~ "Health care services", 43 ~ "Social assistance",
                       44 ~ "Arts/entertainment/recreation", 45 ~ "Traveler accommodation",
                       46 ~ "Food services/drinking places", 47 ~ "Repair/maintenance",
                       48 ~ "Personal/laundry services", 49 ~ "Membership organizations",
                       50 ~ "Private households", 51 ~ "Public administration",
                       .default = "Unknown"
                   ),
                   occupation_code = TRDTOCC1,
                   occupation_desc = case_match(TRDTOCC1,
                       1 ~ "Management", 2 ~ "Business/Finance", 3 ~ "Computer/Math",
                       4 ~ "Architecture/Engineering", 5 ~ "Life/Physical Science",
                       6 ~ "Community/Social Service", 7 ~ "Legal", 8 ~ "Education",
                       9 ~ "Arts/Design/Media", 10 ~ "Healthcare Practitioner",
                       11 ~ "Healthcare Support", 12 ~ "Protective Service",
                       13 ~ "Food Prep/Serving", 14 ~ "Building/Grounds Maintenance",
                       15 ~ "Personal Care", 16 ~ "Sales", 17 ~ "Office/Admin Support",
                       18 ~ "Farming/Fishing/Forestry", 19 ~ "Construction/Extraction",
                       20 ~ "Installation/Repair", 21 ~ "Production",
                       22 ~ "Transportation/Material Moving", .default = "Unknown"
                   ),
                   marital_status = case_match(TRSPPRES,
                       1 ~ "Married, spouse present",
                       2 ~ "Married, spouse absent", 
                       3 ~ "Not married",
                       .default = "Unknown"
                   ),
                   hh_size = TRNUMHOU,
                   college_student = TESCHENR == 1) |> 
            select(matches("[:lower:]", ignore.case = FALSE)), 

        rost = file |> 
            mutate(participant_id = TUCASEID,
                   line_no = TULINENO, # Crucial for identifying the respondent
                   sex = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
                   age = TEAGE,
                   relation_to_resp = TERRP) |>
            select(matches("[:lower:]", ignore.case = FALSE)),
        
   act = file |> 
            mutate(participant_id = TUCASEID,
                   activity_n_of_day = TUACTIVITY_N,
                   time_spent_min = TUACTDUR24,
                   location_code = TEWHERE, 
                   location_desc = case_match(TEWHERE,
                       1 ~ "Home", 2 ~ "Worker's home", 3 ~ "Workplace",
                       4 ~ "Someone else's home", 5 ~ "Restaurant/Bar",
                       6 ~ "Place of worship", 7 ~ "Grocery store",
                       8 ~ "Other store/Mall", 9 ~ "School", 10 ~ "Library",
                       11 ~ "Other place", 12 ~ "Car/Truck/Motorcycle",
                       13 ~ "Bus", 14 ~ "Subway/Train", 15 ~ "Bicycle",
                       16 ~ "Boat/Ferry", 17 ~ "Taxi/Limo",
                       18 ~ "Public transit (Wait/Transfer)", 19 ~ "Airplane",
                       20 ~ "Other mode of transport", 21 ~ "Bank",
                       30 ~ "Gym/Health Club", 31 ~ "Post Office",
                       32 ~ "Government Office",
                       .default = "Unknown/Not reported"
                   ),
                   start_time = hms(TUSTARTTIM),
                   # TIER 1 (2 digits) + 0000
                   level1_code = str_c(str_sub(str_pad(TRTIER2P, 4, pad="0"), 1, 2), "0000"),
                   # TIER 1 (2 digits) + TIER 2 (2 digits) + 00
                   level2_code = str_c(str_pad(TRTIER2P, 4, pad="0"), "00"),
                   level3_code = str_pad(TRCODEP, 6, pad="0")) |>
            select(matches("[:lower:]", ignore.case = FALSE))
    )
    result 
}

# --- FUNCTION: Load Activity Metadata ---
load_atus_activities <- function(){
    dest_file <- file.path("data", "mp02", "atus_activity_codes.csv")
    if(!file.exists(dest_file)){
        download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                      quiet=TRUE, destfile=dest_file, mode="wb")
    }
    # Standardize codes to 6-digit character strings for joining
    read_csv(dest_file, show_col_types = FALSE) |>
        mutate(across(starts_with("code"), ~str_pad(.x, 6, pad="0", side="left")))
}

# --- EXECUTE LOADING ---
resp <- load_atus_data("resp")
rost <- load_atus_data("rost")
act  <- load_atus_data("act")
activities <- load_atus_activities()

# Verify the join will work 
#test_join <- act |> 
#  left_join(activities, by = c("level3_code" = "code_level3")) |> 
#  select(level3_code, task_level3) |> 
#  slice_head(n=5)

#print(test_join)

```



```{r}
# Baseline type/distribution checks
library(dplyr)
library(pillar)  

cat("=== RESPONDENT DATA ===\n")
glimpse(resp); summary(resp); count(resp, survey_year, sort = TRUE)

cat("\n=== ROSTER DATA ===\n") 
glimpse(rost); summary(rost); count(rost, sex, sort = TRUE)

cat("\n=== ACTIVITIES DATA ===\n")
glimpse(act); summary(act)
count(act, activity_n_of_day, sort = TRUE) 

# |> print(n = 20)

#cat("\n=== ACTIVITIES CODES ===\n")
#glimpse(activities); summary(activities)
#count(activities, code_level1, sort = TRUE) 

# |> print(n = 20)
```



**Q1.How many total hours were recorded on sleeping-type activities?**
```{r}
# needs time spent level13 code for sleeping filter and resp for wurvey weight on the populating
sleeping_result <- act |> 
  left_join(resp |> select(participant_id, survey_weight), by = "participant_id") |> 
  filter(str_starts(level3_code, "010101")) |> 
  summarise(total_minutes = sum(time_spent_min * survey_weight, na.rm = TRUE),
            .groups = "drop") 

sleeping_hours <- sleeping_result$total_minutes / 60
scales::comma(sleeping_hours, accuracy = 1, suffix = " hours")
```

**Q2.How many female participants reported spending no time alone in 2003?**
```{r}
#q about the female participants. need resp for usage and  rost to identy who is female
female_alone_2003 <- resp |> 
  filter(survey_year == 2003, time_alone == 0) |> 
  left_join(rost |> filter(sex == "F"), by = "participant_id") |> 
  summarise(n_women_alone = n_distinct(participant_id),
            .groups = "drop")

female_alone_2003

#double check if the above is accurate
# Step 1: 2003 respondents with 0 alone time
resp_2003_alone0 <- resp |> filter(survey_year == 2003, time_alone == 0)
nrow(resp_2003_alone0)  # ~10K total

# Step 2: Of those, how many are female
resp_2003_alone0 |> 
  left_join(rost |> filter(sex == "F"), by = "participant_id") |> 
  nrow()  # Same as distinct count
```


### Exploratory analysis with additional Demographic variables {.tabset .tabset-pills .unnumbered}

### Q1
**How many children under the age of 18 live in the respondent’s home?**
```{r}

library(scales)


kids_summary <- resp |> 
  mutate(has_kids = children_count > 0) |> 
  count(has_kids, name = "HK") |> 
  mutate(
    Count_formatted = scales::comma(HK),
    Status = if_else(has_kids, "Has Children", "No Children")
  ) |> 
  select(Status, Count_formatted)

#kids_summary
```
**A:** `r kids_summary |> filter(Status == "Has Children") |> pull("Count_formatted")` participants have children under 18.

### Q2
**Is the respondent married?**
```{r}
#| echo: false
# 1. Summarize with a safe name 'n'
marital_summary <- resp |> 
  count(marital_status, name = "n") |> 
  # 2. Format the numbers with commas
  mutate(Count = scales::comma(n)) |> 
  # 3. Rename the status column for a clean table
  rename(Status = marital_status) |>
  # 4. Keep only the columns we need
  select(Status, Count)

```
**A:** `r marital_summary |> filter(Status == "Married, spouse present") |> pull("Count")` are married with spouse present.

### Q3 
**What is the respondent’s employment status (employed, unemployed, retired, etc.)?** 
```{r}
# Employment Status Counts - ATUS Respondents
#| echo: false
# 1. Summarize using 'n' to avoid function name conflicts
employment_summary <- resp |> 
  count(employment_status, name = "n") |> 
  # 2. Perform math and formatting using 'n'
  mutate(
    Count = scales::comma(n),
    Pct = scales::percent(n / sum(n), accuracy = 0.1)
  ) |> 
  # 3. Rename for the final table display
  rename(Status = employment_status) |> 
  # 4. Select only the columns for your report
  select(Status, Count, Pct)

#employment_summary
```

**A**: Employment Summary: `r employment_summary |> filter(Status == "Employed - at work") |> pull("Count")` employed, `r employment_summary |> filter(Status != "Employed - at work") |> pull("Count") |> dplyr::first()` not employed.

### Q4
**Location of Activities**
```{r}
location_summary <- act |> 
  # Count occurrences of the physical locations
  count(location_desc, name = "count") |> 
  # Sort by the most frequent locations
  arrange(desc(count)) |> 
  # Take the top 10
  slice_head(n = 10) |> 
  # Formatting for your report
  mutate(
    Count = scales::comma(count),
    `Activity Location` = location_desc
  ) |> 
  select(`Activity Location`, Count)

#location_summary
```

**A:** Top activity location is `r location_summary[["Activity Location"]][1]` with `r location_summary$Count[1]` occurrences.

### Q5
**Is the respondent enrolled in a college or university (not high school)?**
```{r}
#| label: college-university-summary
#| echo: false
#| message: false

library(scales)

# 1. Create the summary table
college_summary <- resp |> 
  count(college_student, name = "n") |> 
  mutate(
    Count_Formatted = scales::comma(n),
    Status = if_else(college_student, "Enrolled in College", "Not Enrolled")
  )

# 2. Calculate raw values for inline text
coll_count_raw <- sum(resp$college_student, na.rm = TRUE)
coll_pct_raw   <- mean(resp$college_student, na.rm = TRUE) * 100

# 3. Pre-format the strings for the A: line
coll_count_f <- scales::comma(coll_count_raw)
coll_pct_f   <- round(coll_pct_raw, 1)
```
**A:** Enrolled in college: `` `r coll_count_f` `` (`` `r coll_pct_f` ``%).

**Data Integration and Initial Exploration - Survey Weighting**  
```{r} 
# All years
resp |> summarise(avg_time_alone = weighted.mean(time_alone, survey_weight, na.rm = TRUE))

# By year
resp |> group_by(survey_year) |> 
  summarise(avg_time_alone = weighted.mean(time_alone, survey_weight, na.rm = TRUE))

# By gender (needs rost)
resp |> 
  left_join(rost |> distinct(participant_id, sex), by = "participant_id") |> 
  distinct(participant_id, sex, time_alone, survey_weight) |> 
  group_by(sex) |> 
  summarise(avg_time_alone = weighted.mean(time_alone, survey_weight, na.rm = TRUE)) 

```

 
### Inline questions {.tabset .tabset-pills .unnumbered}

### Q1
**How many different respondents have answered an ATUS survey since 2003?**
````{r}
#count the respondents 
n_respondents <- nrow(resp)
````

**A:** There are `r scales::comma(n_respondents)` different respondents who have answered an ATUS survey since 2003.


### Q2
**Americans enjoy watching a variety of sports. How many different sports does ATUS ask about watching as a “Level 3” task**

```{r}
#get total of the level 3 answers
sports_count <- act |> 
  filter(str_detect(level3_code, "^1302")) |>  
  distinct(level3_code) |> 
  nrow()
```

**A:** ATUS asks about `r sports_count` different sports as Level 3 watching tasks.

### Q3 
**Approximately what percent of Americans are retired?:**
```{r}
retired_data <- resp |> 
  left_join(rost |> filter(relation_to_resp == 1) |> select(participant_id, age), 
            by = "participant_id") |>  # self only
  mutate(retired = employment_status == "Not in Labor Force" & age >= 65) |> 
  summarise(pct = weighted.mean(retired, survey_weight, na.rm = TRUE) * 100) |> 
  pull(pct) |> round(1)

```

**A:** Approximately `r retired_data`% of Americans are retired.

### Q4
**On average, how many hours do Americans sleep per night?**
```{r}
sleep_hours <- act |> 
  left_join(resp |> select(participant_id, survey_weight), by = "participant_id") |> 
  filter(str_detect(level3_code, "^010")) |> 
  group_by(participant_id) |> 
  summarize(total_sleep_min = sum(time_spent_min, na.rm = TRUE),
            total_weight = first(survey_weight), .groups = "drop") |> 
  summarize(avg_sleep_hours = weighted.mean(total_sleep_min / 60, total_weight, na.rm = TRUE)) |> 
  pull(avg_sleep_hours) |> round(2)

```

**A:** On average, Americans sleep `r sleep_hours` hours per night

### Q5
**On average, how many hours do Americans with one or more children in the home spend taking care of their children?**

```{r}
childcare_hours <- act |> 
  left_join(resp |> select(participant_id, survey_weight, children_count), 
            by = "participant_id") |> 
  filter(children_count > 0, str_detect(level3_code, "^03")) |> 
  group_by(participant_id) |> 
  summarize(total_childcare_min = sum(time_spent_min, na.rm = TRUE),
            total_weight = first(survey_weight), .groups = "drop") |> 
  summarize(avg_childcare_hours = weighted.mean(total_childcare_min / 60, total_weight, na.rm = TRUE)) |> 
  pull(avg_childcare_hours) |> round(2)
```

**A:** Americans with children spend on average `r childcare_hours` hours per day taking care of their children.


## Exploratory Analysis

The following chunks of Code are related to Exploratory Data Analysis assigned by the professor.  This may include exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 


**Q1.Do retired persons spend more time on lawn-care than people who have full-time employment?**

```{r}
#| label: tbl-lawn-retired
#| tbl-cap: "Lawn Care Hours: Retired vs Full-Time Employed"
lawn_care <- resp |> 
  #  Join to rost using line_no = 1 to get the respondent's age correctly
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age), 
            by = "participant_id") |> 
  #  Use the specific Level 3 code for Lawn & Garden (020501)
  left_join(act |> filter(level3_code == "020501") |> 
            group_by(participant_id) |> 
            summarise(total_lawn_min = sum(time_spent_min, na.rm = TRUE), .groups = "drop"),
            by = "participant_id") |> 
  mutate(total_lawn_min = replace_na(total_lawn_min, 0),
         employment_group = case_when(
           employment_status == "Employed" ~ "Full-time Employed",
           employment_status == "Not in Labor Force" & age >= 65 ~ "Retired",
           TRUE ~ NA_character_
         )) |> 
  filter(!is.na(employment_group)) |> 
  group_by(employment_group) |> 
  summarise(avg_lawn_hours = weighted.mean(total_lawn_min/60, survey_weight, na.rm = TRUE),
            .groups = "drop")

# Table
lawn_care |> 
  gt() |> 
  fmt_number(columns = avg_lawn_hours, decimals = 3) |> 
  cols_label(employment_group = "Employment Status", avg_lawn_hours = "Avg Hours/Day") |> 
  tab_header("Lawn Care Time by Employment Status")

```
**A:** On Average, retired folks do spend more time on lawn care than full-time employed people.The difference is 0.209 hours per day, or about 12.5 minutes, which likely reflects the fact that most retirees have more free time for home and yard upkeep.


**Q2.How does the amount of time spent on social and recreational activities typically compare for married and unmarried people?**


```{r}
#| label: tbl-social-marital
#| tbl-cap: "Social/Recreation Hours: Married vs Unmarried"
social_rec <- act |> 
  left_join(resp |> select(participant_id, marital_status, survey_weight), 
            by = "participant_id") |> 
  filter(str_detect(level3_code, "^(12|13)")) |> 
  # STEP 1: Get total rec minutes PER PERSON
  group_by(participant_id, marital_status) |> 
  summarise(total_rec_min = sum(time_spent_min, na.rm = TRUE),
            survey_weight = first(survey_weight),  # Keep ONE weight per person
            .groups = "drop") |> 
  # STEP 2: Create simple groups
  mutate(marital_group = case_when(
    str_detect(marital_status, "Married") ~ "Married",
    TRUE ~ "Unmarried"
  )) |> 
  # STEP 3: ONE weighted average per group
  group_by(marital_group) |> 
  summarise(avg_rec_hours = weighted.mean(total_rec_min/60, survey_weight, na.rm = TRUE),
            .groups = "drop")

social_rec |> 
  gt() |> 
  fmt_number(columns = avg_rec_hours, decimals = 2) |> 
  cols_label(
    marital_group = "Marital Status",
    avg_rec_hours = "Avg Hours/Day"
  ) |> 
  tab_header("Social & Recreation Time by Marital Status")


```
**A:** As expected, individuals who are married have less time for social and recreation than their unmarried peers. 


**Q3.Where do high-earning Americans spend the majority of their time?**

```{r}
#| label: tbl-high-earners
#| tbl-cap: "High-Earners (>90th percentile wage) Time Allocation"

# 1. Define the threshold
high_wage_cut <- quantile(resp$hourly_wage[resp$hourly_wage > 0], 0.9, na.rm = TRUE)

# 2. Process the data
high_earners_clean <- act |> 
  left_join(resp |> select(participant_id, hourly_wage, survey_weight), 
            by = "participant_id") |> 
  filter(!is.na(hourly_wage), hourly_wage >= high_wage_cut) |> 
  left_join(activities |> distinct(code_level3, task_level1), 
            by = c("level3_code" = "code_level3")) |> 
  mutate(activity_group = coalesce(task_level1, "Other/Unknown")) |> 
  group_by(activity_group) |> 
  # Use weights for the calculation
  summarise(total_weighted_time = sum(time_spent_min * survey_weight, na.rm = TRUE),
            .groups = "drop") |> 
  # Convert to a percentage
  mutate(pct_time = total_weighted_time / sum(total_weighted_time)) |> 
  arrange(desc(pct_time)) |> 
  slice_head(n = 5) |>
  # REMOVE THE WEIGHTS: Only keep the display columns
  select(activity_group, pct_time)

# 3. Display with gt
high_earners_clean |> 
  gt() |> 
  fmt_percent(columns = pct_time, decimals = 1) |> 
  cols_label(
    activity_group = "Activity Category", 
    pct_time = "% of Daily Time"
  ) |> 
  tab_header(
    title = "How High Earners Use Their Day",
    subtitle = glue("Top 10% Earners (Wages >= ${round(high_wage_cut, 2)}/hr)")
  )
```
**A:** High-earning American (top 10% hourly wage), spend the majority of their time in personal care activities(~38%), followed by by work (~21%), and socializing, relaxing, and leisure activities. 


**Q4. What are the most common activities reported by Americans when on a bus, train, subway, boat, or ferry? **
```{r}
#| label: tbl-transit
#| tbl-cap: "Top Activities While Using Public Transit"

transit_activities <- act |> 
  # Filter by the physical locations mapped earlier
  filter(location_desc %in% c("Bus", "Subway/Train", "Boat/Ferry")) |> 
  # Join to get the Activity names (Level 3)
  left_join(activities |> select(code_level3, task_level3), 
            by = c("level3_code" = "code_level3")) |> 
  # Group by the specific activity being done
  count(task_level3, name = "Occurrences") |> 
  filter(!is.na(task_level3)) |> 
  arrange(desc(Occurrences)) |> 
  slice_head(n = 5) |> 
  # Format for the table
  mutate(Occurrences = scales::comma(Occurrences))

# Display Table
transit_activities |> 
  gt() |> 
  cols_label(task_level3 = "Activity Being Performed", Occurrences = "Count") |> 
  tab_header(
    title = "What are Americans doing on Public Transit?",
    subtitle = "Data filtered for Bus, Subway, Train, and Ferry locations"
  )
```

**A:** The most common activities for Americans while on a bus, train, subway, boat, or ferry beside traveling, are socializing, eating and drivingking, and unfortunately working. 




**Q5.How much time per day do Americans spend on your four favorite activities and how does this vary by age group?**
```{r}
#| label: tbl-miguel-age-analysis
#| tbl-cap: "Life Stage Analysis: How Time Use Shifts Across Generations"

# 1. High-Contrast Activity Codes
# 060100 = Education, 120100 = Socializing, 150100 = Volunteering, 050100 = Working
target_codes <- c("060100", "120100", "150100", "050100")

# 2. The Final Weighted Summary
age_activity_summary <- act |> 
  filter(level2_code %in% target_codes) |> 
  left_join(resp |> select(participant_id, survey_weight), by = "participant_id") |> 
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age), by = "participant_id") |> 
  left_join(activities |> 
              mutate(code_level2 = str_pad(as.character(code_level2), 6, pad="0")) |>
              distinct(code_level2, task_level2), 
            by = c("level2_code" = "code_level2")) |> 
  # Using cut() to 'bin' age into the professor's suggested buckets
  mutate(age_group = cut(age, 
                         breaks = c(18, 35, 50, 65, Inf), 
                         labels = c("18-34", "35-49", "50-64", "65+"), 
                         right = FALSE)) |> 
  filter(!is.na(age_group)) |> 
  group_by(task_level2, age_group) |> 
  summarise(
    avg_hours = sum(time_spent_min * survey_weight, na.rm = TRUE) / 
                sum(survey_weight, na.rm = TRUE) / 60,
    .groups = "drop"
  ) |> 
  pivot_wider(names_from = age_group, values_from = avg_hours) |> 
  rename(Activity = task_level2)

# 3. Create the GT Table
age_activity_summary |> 
  gt() |> 
  fmt_number(columns = -Activity, decimals = 2) |> 
  cols_label(Activity = "Activity Category") |> 
  tab_header(
    title = "How Do Americans Spend Their Day?",
    subtitle = "Average daily hours by life stage (Weighted Analysis)"
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Activity)
  )
```
**A:** Based on the data, taking a Class and Working show the most significant decline as Americans age, dropping from 2.89 to 1.72 hours and 3.22 to 2.76 hours, respectively.

Conversely, Socializing and Administrative Activities remain remarkably stable across all life stages, peaking slightly for those aged 65 and older.

## Task 6: Exploratory Questions - Graph Answers

**Q1 How does the amount of time spent on household activities change as Americans age?**
```{r}
#| label: fig-household-age
#| fig-cap: "Average Daily Time on Household Activities by Age Group"

q1_data <- act |>
  filter(level1_code == "020000") |>
  left_join(resp |> select(participant_id, survey_weight), by = "participant_id") |>
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age), by = "participant_id") |>
  filter(!is.na(age)) |>
  mutate(age_bin = cut(age, breaks = seq(15, 85, by = 5), right = FALSE)) |>
  group_by(age_bin) |>
  summarize(avg_hours = sum(time_spent_min * survey_weight) / sum(survey_weight) / 60) |>
  filter(!is.na(age_bin))

ggplot(q1_data, aes(x = age_bin, y = avg_hours, group = 1)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal() +
  labs(x = "Age Group", y = "Average Hours per Day") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**A:**
Time spent on household activities such as cleaning, cooking, and maintenance, shows a steady increase throughout a person's lifespan, peaking significantly in the retirement years (age 65+). Young adults (18-24) spend less than an hour on these tasks while retirees often dedicate nearly 2.5 hours daily, suggesting that household management becomes a primary "filler" of time once one retires from the work force. 

**Q2.What are the most popular activities for Americans employed in education during the summer months (June, July, and August)**
```{r}
#| label: fig-summer-edu-fixed
#| fig-cap: "Top 10 Summer Activities for Education Sector Employees (Excluding Work)"

# 1. Process the data
q2_data <- resp |>
  filter(industry_desc == "Educational services",
         diary_month %in% c("Jun", "Jul", "Aug")) |>
  select(participant_id, survey_weight) |>
  inner_join(act, by = "participant_id") |>
  left_join(activities |> distinct(code_level2, task_level2), by = c("level2_code" = "code_level2")) |>
  # Exclude Working (05) but keep Sleep (01)
  filter(!level1_code %in% c("050000")) |> 
  group_by(task_level2) |>
  summarise(total_weighted_min = sum(time_spent_min * survey_weight, na.rm = TRUE), .groups = "drop") |>
  mutate(avg_daily_min = total_weighted_min / sum(resp$survey_weight[resp$industry_desc == "Educational services" & resp$diary_month %in% c("Jun", "Jul", "Aug")])) |>
  arrange(desc(avg_daily_min)) |>
  slice_head(n = 10) |>
  # Wrap the activity names so they don't overlap or cut off on the Y-axis
  mutate(task_level2 = stringr::str_wrap(task_level2, width = 20))

# 2. Plot with Title Wrapping
ggplot(q2_data, aes(x = reorder(task_level2, avg_daily_min), y = avg_daily_min/60)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  theme_minimal() +
  labs(
    # Wrap the title at 40 characters to prevent it from bleeding off the edge
    title = stringr::str_wrap("How Educators Spend Summer Days (Work-Free View)", width = 40),
    subtitle = "Top activities for Educational Services employees in Jun-Aug",
    x = NULL, 
    y = "Average Hours per Day"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 9)
  )
```
**A:** 
Summer data for education professionals reveals a clear 'recharge' phase. With work removed, 'Relaxing and Leisure' takes center stage at 3 hours daily, while 'Socializing' and 'Eating and Drinking' fill the remaining free time. This highlights a seasonal transition where educators prioritize personal recovery and social connection over professional or academic structure.

**Q3.How does the amount of time allocated to various educational activities (attending class, doing homework, studying, etc.) vary by age?**
```{r}
#| label: fig-edu-by-age-clean
#| fig-cap: "How Students Allocate Their Study Time: Formal Class vs. Independent Work"

q3_clean <- resp |>
  filter(college_student == TRUE) |>
  inner_join(act |> filter(str_starts(level2_code, "06")), by = "participant_id") |>
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age), by = "participant_id") |>
  # 1. Simplify the 10+ categories into 3 meaningful ones
  mutate(task_simplified = case_when(
    str_detect(level3_code, "060101|060102") ~ "Attending Class",
    str_detect(level3_code, "0603") ~ "Homework & Research",
    TRUE ~ "Other (Admin/Extra)"
  )) |>
  mutate(age_group = cut(age, breaks = c(18, 25, 35, 50), 
                         labels = c("18-24", "25-34", "35-49"), right = FALSE)) |>
  filter(!is.na(age_group)) |>
  group_by(age_group, task_simplified) |>
  summarize(total_min = sum(time_spent_min * survey_weight), .groups = "drop_last") |>
  # 2. Calculate the % of total education time for that age group
  mutate(pct_time = total_min / sum(total_min))

# 3. Plot a "Filled" bar chart to show the shift in focus
ggplot(q3_clean, aes(x = age_group, y = pct_time, fill = task_simplified)) +
  geom_col(position = "fill", width = 0.7) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Focus Shifts from Class to Homework as Students Age",
    subtitle = "Percentage of total education time by age group",
    x = "Age Group", y = "Share of Study Time", fill = "Activity"
  ) +
  theme(legend.position = "top")
```
**A:** 
While younger students (18-24) spend a majority of their time physically attending class, older students (35-49) shift their focus significantly toward independent homework and research. This suggests that while "education is wasted on the young" might be an exaggeration, older students do show a more self-directed and intensive approach to their studies.

**Q4.Show how the relative interest in different sports changes as a function of either age.**
```{r}
#| label: fig-sports-bump
#| fig-cap: "Ranking of Sport Popularity by Age Group"

q4_data <- act |>
  filter(str_starts(level2_code, "1301")) |>
  left_join(resp |> select(participant_id, survey_weight), by = "participant_id") |>
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age), by = "participant_id") |>
  left_join(activities |> distinct(code_level3, task_level3), by = c("level3_code" = "code_level3")) |>
  mutate(age_group = cut(age, breaks = c(15, 25, 35, 45, 55, 65, Inf), 
                         labels = c("15-24", "25-34", "35-44", "45-54", "55-64", "65+"), right = FALSE)) |>
  filter(!is.na(age_group)) |>
  group_by(age_group, task_level3) |>
  summarize(total_time = sum(time_spent_min * survey_weight), .groups = "drop") |>
  group_by(age_group) |>
  mutate(rank = rank(-total_time, ties.method = "random")) |>
  filter(rank <= 5) |>
  ungroup()

ggplot(q4_data, aes(x = age_group, y = rank, color = task_level3, group = task_level3)) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  scale_y_reverse(breaks = 1:5) +
  theme_minimal() +
  labs(x = "Age Group", y = "Rank (1 = Most Popular)", color = "Sport")
```
**A:**
The "Bump Plot" reveals a significant transition in physical activity as Americans age. High-intensity team sports and "Walking" dominate the early years, but as users enter their 40s and 50s, individualistic and lower-impact activities like "Golf" and "Weightlifting/Strength Training" climb the rankings.

**Q5** Middle-child syndrome
```{r}
#| label: fig-childcare-per-child
#| fig-cap: "The 'Dilution' of Parental Attention: Daily Care Minutes Per Child"

# 1. Calculate total childcare time per person
# We use Level 2 codes starting with 0301, 0302, 0303 (Caring for HH children)
childcare_summary <- act |>
  filter(str_starts(level2_code, "0301|0302|0303")) |>
  group_by(participant_id) |>
  summarize(total_child_min = sum(time_spent_min), .groups = "drop")

# 2. Join with respondent data to get child counts and weights
childcare_analysis <- resp |>
  select(participant_id, children_count, survey_weight) |>
  filter(children_count > 0) |> # Only include households with children
  left_join(childcare_summary, by = "participant_id") |>
  # Replace NA with 0 for parents who didn't record childcare on their diary day
  mutate(total_child_min = replace_na(total_child_min, 0)) |>
  # Calculate the primary metric: Time Per Child
  mutate(time_per_child = total_child_min / children_count)

# 3. Group by Family Size and calculate weighted averages
# We group 5+ children together to ensure a healthy sample size
plot_data <- childcare_analysis |>
  mutate(family_size = ifelse(children_count >= 5, "5+", as.character(children_count))) |>
  group_by(family_size) |>
  summarize(
    avg_min_per_child = sum(time_per_child * survey_weight) / sum(survey_weight),
    .groups = "drop"
  ) |>
  mutate(family_size = factor(family_size, levels = c("1", "2", "3", "4", "5+")))

# 4. Create a publication-quality bar chart
ggplot(plot_data, aes(x = family_size, y = avg_min_per_child)) +
  geom_col(fill = "#4E79A7", width = 0.7) +
  geom_text(aes(label = round(avg_min_per_child, 0)), vjust = -0.5, fontface = "bold") +
  theme_minimal() +
  labs(
    title = "Does More Children Mean Less Attention?",
    subtitle = "Average daily childcare minutes allocated per child by family size",
    x = "Number of Children in Household",
    y = "Average Minutes per Child"
  ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(face = "italic")
  )

```
**A:**
As family size increases, the amount of direct parental attention each child receives drops sharply, falling from about 140 minutes in one-child households to just 55 minutes in households with four children. Although total childcare time rises somewhat as families grow, it does not increase at the same pace, suggesting that parental time has to be spread more thinly across more children.

This also helps explain the “middle-child” experience: in larger families, each child tends to receive a smaller share of parental engagement than children in smaller families



# Code for Final insights an deliverable

Code related to the final deliverable of the assignment goes here. 
```{r}
#| label: market-research-analysis-final

# 1. Identify Target Groups using your CLEANED variable names
# Demographic 3: "Young Professionals" (Our custom choice)
market_groups <- resp |>
  # Join with roster to get age/sex
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age, sex), by = "participant_id") |>
  mutate(target_group = case_when(
    # SAHDs: Male, Has children, Not in labor force
    sex == "M" & children_count > 0 & employment_status == "Not in Labor Force" ~ "Stay-at-Home Dads",
    
    # Undergrads: 18-24 and college_student flag is TRUE
    age >= 18 & age <= 24 & college_student == TRUE ~ "College Undergrads",
    
    # Young Professionals: 22-29, Employed, No children
    age >= 22 & age <= 29 & employment_status == "Employed" & children_count == 0 ~ "Young Professionals",
    
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(target_group)) |>
  select(participant_id, target_group, survey_weight)

# 2. Join with Activity Data and Calculate Percentiles
market_results <- act |>
  inner_join(market_groups, by = "participant_id") |>
  # Grouping Level 1 codes into app-relevant categories
  mutate(activity_name = case_when(
    str_starts(level1_code, "01") ~ "Sleeping",
    str_starts(level2_code, "0301") ~ "Direct Childcare",
    str_starts(level1_code, "06") ~ "Education/Studying",
    str_starts(level1_code, "13") ~ "Exercise",
    TRUE ~ "Other"
  )) |>
  filter(activity_name != "Other") |>
  group_by(target_group, activity_name) |>
  summarise(
    # Percentiles represent Low, Typical, and High users
    low = quantile(time_spent_min / 60, 0.25, na.rm = TRUE),
    typical = quantile(time_spent_min / 60, 0.50, na.rm = TRUE),
    high = quantile(time_spent_min / 60, 0.75, na.rm = TRUE),
    # App Target Level set at the 85th percentile (top performers)
    target_level = quantile(time_spent_min / 60, 0.85, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Final Step: Calculate fraction of people hitting the target
# This creates the final metric for your pitch
final_market_table <- market_results |>
  # Cross-reference back to individual data to see who beats the target_level
  left_join(
    act |> 
      inner_join(market_groups, by = "participant_id") |>
      mutate(activity_name = case_when(
        str_starts(level1_code, "01") ~ "Sleeping",
        str_starts(level2_code, "0301") ~ "Direct Childcare",
        str_starts(level1_code, "06") ~ "Education/Studying",
        str_starts(level1_code, "13") ~ "Exercise",
        TRUE ~ "Other"
      )) |>
      group_by(target_group, participant_id, activity_name) |>
      summarise(indiv_hrs = sum(time_spent_min)/60, .groups = "drop"),
    by = c("target_group", "activity_name")
  ) |>
  group_by(target_group, activity_name) |>
  summarise(
    low = first(low), 
    typical = first(typical), 
    high = first(high), 
    target = first(target_level),
    pct_hitting_target = mean(indiv_hrs >= target_level) * 100,
    .groups = "drop"
  )

# 4. Display the Market Research Table
final_market_table |>
  gt() |>
  tab_header(title = "Time-Tracking App: Target User Benchmarks",
             subtitle = "Estimates for 'Typical' and 'Target' time use patterns") |>
  fmt_number(columns = c(low, typical, high, target), decimals = 2) |>
  fmt_number(columns = pct_hitting_target, decimals = 1, suffix = "%") |>
  cols_label(
    target_group = "Demographic",
    activity_name = "Activity",
    pct_hitting_target = "% Meeting Target"
  )

```

## Final Insights and Deliverable

Our time-tracking app moves beyond simple scheduling by leveraging Social Comparison Theory. By providing users with "Peer Benchmarks," we transform abstract time data into actionable, competitive, and rewarding goals.


**Target Demographic Definitions**
To power our gamification engine, we have identified three high-value segments based on life-stage and time-use priorities:

**Stay-at-Home Dads (SAHDs)**: Defined as males with at least one child in the household who are not currently in the labor force.

**College Undergrads**: Students aged 18–24 currently enrolled in a degree program.

**Young Professionals (The "Hustle" Segment)**: Employed adults aged 22–29 with no children, focused on fitness and career.

**Activity Benchmarking**
We analyzed 20+ years of ATUS data to establish what "Typical" looks like versus an "Aspirational Target" (defined as the 85th percentile of peers).

**Strategic Insights for the UI/UX Team: The Gamification "Nudge"**
The visualization below reveals clear, impactful "gaps to greatness" across all three of our target segments. We can use these benchmarks to trigger relevant notifications and reward milestones, moving users from "Typical" to "Top Tier."


```{r}
#| label: fig-market-research-with-legend
#| fig-cap: "Daily Time Benchmarks: Typical vs. Target Performance"

# 1. GENERATE THE DATA (This ensures the object exists when the plot runs)
market_plot_data <- resp |>
  left_join(rost |> filter(line_no == 1) |> select(participant_id, age, sex), by = "participant_id") |>
  mutate(target_group = case_when(
    sex == "M" & children_count > 0 & employment_status == "Not in Labor Force" ~ "Stay-at-Home Dads",
    age >= 18 & age <= 24 & college_student == TRUE ~ "College Undergrads",
    age >= 22 & age <= 29 & employment_status == "Employed" & children_count == 0 ~ "Young Professionals",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(target_group)) |>
  inner_join(act, by = "participant_id") |>
  mutate(activity_name = case_when(
    str_starts(level1_code, "01") ~ "Sleeping",
    str_starts(level2_code, "0301") ~ "Childcare",
    str_starts(level1_code, "06") ~ "Education",
    str_starts(level1_code, "13") ~ "Exercise",
    TRUE ~ "Other"
  )) |>
  filter(activity_name != "Other") |>
  group_by(target_group, activity_name) |>
  summarise(
    typical = quantile(time_spent_min / 60, 0.50, na.rm = TRUE),
    target = quantile(time_spent_min / 60, 0.85, na.rm = TRUE),
    .groups = "drop"
  )

# 2. CREATE THE PLOT
ggplot(market_plot_data, aes(x = activity_name, y = typical, fill = activity_name)) +
  geom_col(alpha = 0.7, width = 0.6) +
  geom_errorbar(aes(ymin = target, ymax = target, linetype = "App Target (85th Pct)"), 
                color = "red", linewidth = 1) +
  facet_wrap(~target_group, scales = "free_x") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  scale_linetype_manual(name = "Benchmarks", values = "dashed") +
  labs(
    title = "Typical vs. Target Daily Time Use",
    subtitle = "Comparing median behavior against top-tier performance goals",
    x = NULL, 
    y = "Daily Hours",
    fill = "Activity Type"
  ) +
  theme(
    legend.box = "vertical", 
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  guides(fill = guide_legend(nrow = 1, order = 1), 
         linetype = guide_legend(order = 2))
```

:::{.callout-note title="Stay-at-Home Dads: The 'Deep Connection' Badge"}
The data shows a massive opportunity here. The typical dad's direct, focused engagement with his children is relatively low (just under 1 hour). Yet, our App Target (85th Pct) is over 4 hours. This huge discrepancy allows us to create an elite milestone that gamifies deep parental engagement. We will launch a "Quality Time" badge for anyone who can bridge this 3+ hour gap.
:::

:::{.callout-tip title="College Undergrads: The 'Sleep Hygeine' Streak"}
This is our biggest potential win for user health. The data reveals a massive "Sleep Gap"—nearly 4 hours daily between the typical student and our top performers. Our app must prioritize sleep gamification. We will introduce a "Healthy Body, Healthy Mind" streak that rewards consistent, 85th-percentile sleep habits, particularly during stressful midterm and final exam weeks.
:::

:::{.callout-warning title="Young Professionals: Breaking the 'Exercise Wall'"}
For our "Hustle" segment, the typical exercise time is low—closer to the zero mark than we would like to see. To keep them motivated, we will not lead with the 1.5-hour "App Target" immediately. Instead, our UI will offer "Rookie of the Year" milestones that celebrate even 20-30 minutes of daily activity as being "Better than Average" among their peers. We will reserve the full 85th-percentile benchmark for a final "Iron Peak" challenge.


## Extra credit
```{r}
#| label: fig-productivity-rings
#| message: false
#| warning: false

library(tidyverse)

# 1. Data Prep
plot_data <- tribble(
    ~Activity,   ~Target, ~Actual,
    "Exercise",  1.8,     1.2,
    "Education", 3.5,     4.2, 
    "Sleep",     8.5,     6.0  
) %>%
    mutate(
        fraction = Actual / Target,
        # Split the progress into two parts: the first 100% and the "Extra"
        base_progress = pmin(fraction, 1),
        extra_progress = pmax(0, fraction - 1),
        ring_pos = case_when(
            Activity == "Exercise" ~ 1,
            Activity == "Education" ~ 2,
            Activity == "Sleep" ~ 3
        )
    )

# 2. Plot
ggplot(plot_data, aes(x = ring_pos, fill = Activity)) +
    # Background Tracks (Dark Gray)
    geom_col(aes(y = 1), fill = "#2c2c2e", width = 0.8) +
    
    # Base Progress (Up to 100%)
    geom_col(aes(y = base_progress), width = 0.8, show.legend = FALSE) +
    
    # Extra Progress (The "Overlap" for Education)
    # We use a slightly different alpha or color to make it 'pop'
    geom_col(aes(y = extra_progress), width = 0.8, show.legend = FALSE, alpha = 0.8) +

    coord_polar(theta = "y") +
    scale_x_continuous(limits = c(0, 4)) + 
    scale_y_continuous(limits = c(0, 1)) + 
    scale_fill_manual(values = c(
        "Exercise" = "#fa114f", 
        "Education" = "#9efd38", 
        "Sleep" = "#00d2ff"
    )) +
    theme_void() +
    labs(
        title = "Weekly Intensity",
        subtitle = "Education (Green) is crushing it at 120%!",
        caption = "Red: Exercise | Green: Education | Blue: Sleep"
    ) +
    theme(
        plot.background = element_rect(fill = "black", color = "black"),
        panel.background = element_rect(fill = "black", color = "black"),
        plot.title = element_text(color = "white", face = "bold", size = 22, hjust = 0.5),
        plot.subtitle = element_text(color = "#9efd38", size = 12, hjust = 0.5),
        plot.caption = element_text(color = "#8e8e93", size = 10, hjust = 0.5)
    )
```


::: {.callout-note title="AI Usage Statement" collapse="true"}

Perplexity was used to help write the code in this project, but all non-code text was generated without the use of any Generative AI tools. Additionally, Perplexity was used to provide additional background information on the topic and to brainstorm ideas for the final open-ended prompt.

:::

------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
