---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "miggonzalez"
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

Every decade the U.S. Census Bureau redraws the political map through congressional reapportionment, shifting House seats among the states based on how populations change. Most of the discussion focuses on births, deaths and international migration, but internal migration the movement of people across state lines quietly shapes these outcomes just as much.

In this mini project I use American Community Survey (ACS) 1 year data to estimate where Americans are moving within the country. From those migration flows I build a simple demographic model that projects state level populations over the next ten years and explores how those changes might translate into gains or losses of congressional seats. The aim is to connect the everyday decisions of people moving from one state to another to the bigger picture of how political power shifts across the U.S.

### Data Acquisition {-}
::: {.panel-tabset .nav-pills}
### State Population and Birth Information
```{r}
library(tidyverse)
library(tidycensus)

# 1. Set up local cache
data_dir <- "data/mp03_census_cache"
if(!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

options(tidy_census_cache_path = data_dir)    # where tidycensus saves data
options(tigris_use_cache     = TRUE)          # cache shapefiles if you add them later

# 2. Tidy‑style function for one year (no geometry for now)
get_state_pop_year <- function(year) {
  dat <- get_acs(
    geography   = "state",
    variables   = c(total_population = "B01003_001"),
    year        = year,
    survey      = "acs1",
    geometry    = FALSE,
    cache_table = TRUE
  )
  
  dat %>%
    filter(variable == "total_population") %>%
    select(NAME, estimate) %>% 
    rename(
      state_name = NAME,
      total_population = estimate
    ) %>%
    mutate(year = year)      # add the year from the function argument
}

# 3. Run over 2015–2024
YEARS <- c(2015:2019, 2021:2024)

state_pop_df <- map_dfr(YEARS, get_state_pop_year)

# 4. Check that it worked
#state_pop_df
```
**Please note**
This table represents the baseline state population tables which contains the total population estimates for each state and U.S territories.Due to the non-release of the 2020 1-year ACS report, the baseline population tables **excludes 2020** and instead uses the ACS-1 estimates for 2015-2019 and 2021-2024.


### 2024 migration data aquisition 
```{r}
library(tidyverse)
library(readxl)

data_dir_mp03 <- "data/mp03"
if(!dir.exists(data_dir_mp03)) dir.create(data_dir_mp03, recursive = TRUE)

url_2024 <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
file_2024 <- file.path(data_dir_mp03, "State_to_State_Migration_Table_2024_T13.xlsx")

state_to_state_migration_2024 <- function() {
  # Download if missing
  if(!file.exists(file_2024)) {
    message("Downloading 2024 migration table to ", file_2024)
    download.file(url_2024, destfile = file_2024, mode = "wb")
  }
  
  # Read and clean the "Table" sheet (moves between states)
  tab <- read_excel(
    file_2024,
    sheet     = "Table",
    skip      = 13,          # adjust based on your file
    col_names = FALSE
  )
  
  # Clean and tidy the main migration table
  tab_tidy <- tab %>%
    # DO NOT include this line:
    # select(!is.na()) %>%     # drop empty rows if needed
    mutate(across(everything(), ~replace_na(as.character(.), "0"))) %>%
    pivot_longer(
      cols = -1,
      names_to = "state_1y",
      values_to = "population"
    ) %>%
    # Turn "N" into "0" to avoid NAs on coercion
    mutate(
      population = ifelse(population == "N", "0", population)
    ) %>%
    # Convert to integer; warning "NAs introduced by coercion" should now vanish
    mutate(population = as.integer(population)) %>%
    # --- ADD THIS: turn ...3, ...4 into real state names/abbreviations ---
    # In practice, you’ll want a full lookup table, but this is a minimal example:
    mutate(
      state_1y = case_when(
        state_1y == "...3" ~ "ALABAMA",
        state_1y == "...4" ~ "ALASKA",
        # Add others as needed, or use a lookup table
        TRUE ~ "UNKNOWN"
      )
    ) %>%
    # Create state_current from the first column of the original data
    mutate(
      state_current  = toupper(.[[1]]),
      state_1y       = toupper(state_1y),
      year_current   = 2024,
      year_1y        = 2023
    ) %>%
    select(state_current, state_1y, population, year_current, year_1y) %>%
    filter(population > 0)   # remove "non‑moving" rows with 0 or NA
  
  # Read and clean the "Supplemental - Current Res" sheet (stationary)
  suppl <- read_excel(
    file_2024,
    sheet     = "Supplemental - Current Res",
    skip      = 7,
    col_names = FALSE
  )
  
  suppl_stay <- suppl %>%
    mutate(
      state_current  = toupper(.[[1]]),
      state_1y       = toupper(.[[1]]),
      population     = as.integer(.[[2]]),  # pick the right column
      year_current   = 2024,
      year_1y        = 2023
    ) %>%
    select(state_current, state_1y, population, year_current, year_1y)
  
  # Combine both parts
  bind_rows(tab_tidy, suppl_stay)
}
mig_2024 <- state_to_state_migration_2024()
```
I stored the 2024 state-to-state migration flows in mig_2024, with columns state_current, state_1y, population, year_current = 2024, and year_1y = 2023. Each row represents the number of people who moved or remained between two states from 2023 to 2024.

### 2023 migration data aquisition 

```{r task3, cache=TRUE}
library(tidyverse)
library(readxl)

file_path <- "data/mp03/State_to_State_Migration_Table_2023_T13.xlsx"
states <- toupper(state.name)

# 1. STAYERS - Clean extraction
stayers_raw <- read_excel(file_path, range = "B9:B58", col_names = FALSE)
stayers <- tibble(
  state_current = states,
  state_1y = states,
  population = as.integer(pull(stayers_raw, 1)),
  year_current = 2023,
  year_1y = 2022
)

# 2. MOVERS - CRITICAL: Force ALL columns to character FIRST
raw_data <- read_excel(file_path, skip = 8, col_names = FALSE)
colnames(raw_data) <- paste0("col", 1:ncol(raw_data))

mover_data <- raw_data %>%
  slice(2:51) %>%  # 50 state rows
  select(col2:col51) %>%  # 50 state columns
  mutate(across(everything(), as.character)) %>%  # FIX: ALL character!
  pivot_longer(everything(), 
               names_to = "col_name", 
               values_to = "population_raw")

# 3. Add state identifiers
mover_long <- mover_data %>%
  mutate(
    col_id = as.integer(str_remove(col_name, "col")),
    row_id = rep(1:50, each = 50),
    state_current = states[row_id],
    state_1y = states[col_id],
    population = case_when(
      population_raw %in% c("N", NA_character_, "") ~ NA_integer_,
      TRUE ~ as.integer(population_raw)
    ),
    year_current = 2023,
    year_1y = 2022
  ) %>%
  filter(!is.na(population), population > 0, state_current != state_1y) %>%
  select(state_current, state_1y, population, year_current, year_1y)

# 4. FINAL COMBINE
migration_2023 <- bind_rows(stayers, mover_long)

# 5. VERIFICATION
cat("Rows:", nrow(migration_2023), "Columns:", ncol(migration_2023), "\n")
cat("Stayers:", sum(migration_2023$state_current == migration_2023$state_1y), "\n")
#head(migration_2023, 10)
```

I stored the 2023 state-to-state migration flows in mig_2023, with columns state_current, state_1y, population, year_current = 2023, and year_1y = 2022. Each row represents the number of people who moved or remained between two states from 2022 to 2023.

### Combining 2023 and 2024 data 
```{r}
# Task 4: Combine your 2023 + 2024 functions
state_to_state_migration <- function() {
  
  # Call YOUR 2024 function (already working)
  mig_2024 <- state_to_state_migration_2024()
  
  # Call YOUR 2023 function (already working as migration_2023)
  mig_2023 <- migration_2023  # From your Task 3 chunk
  
  # COMBINE with bind_rows() → ~5,000 rows total
  combined_migration <- bind_rows(mig_2024, mig_2023)
  
  return(combined_migration)
}

# TEST THE FUNCTION
migration_combined <- state_to_state_migration()

# VERIFICATION (matches professor's requirements)
cat("Rows:", nrow(migration_combined), "Columns:", ncol(migration_combined), "\n")
cat("Years:", paste(unique(migration_combined$year_current), collapse = ", "), "\n")
cat("Expected: ~5,000 rows x 5 columns\n")
#head(migration_combined, 12)
```


### Metro-to-Metro Migration Flows

```{r task5}
library(tidyverse)
library(jsonlite)
library(stringr)

move_row_to_colnames <- function(X, row = 1){
  X <- as_tibble(X)
  Xrow <- X[row,]
  X <- X[-row,]
  colnames(X) <- Xrow
  readr::type_convert(X)
}

pull_state_from_metro <- function(metro_name){
  str_extract(metro_name, ".*, (\\S{2})[-[:alpha:]]* Metro Area", group=1)
}

# MAIN FUNCTION: Metro-to-metro flows from Census API
metro_to_metro_migration <- function() {
  
  # 1. Create directory
  data_dir <- "data/mp03"
  if(!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)
  
  # 2. Check for existing JSON file
  json_file <- file.path(data_dir, "metro_to_metro_migration_2020.json")
  
  # 3. API URL - 2020 ACS Flows, MSA/Micropolitan level
  api_url <- "https://api.census.gov/data/2020/acs/flows?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
  
  # 4. Download if missing
  if(!file.exists(json_file)) {
    message("Downloading metro-to-metro flows from Census API...")
    download.file(api_url, destfile = json_file)
  }
  
  # 5. Read JSON and fix header row
  metro_raw <- read_json(json_file, simplifyVector = TRUE)
  metro_tidy <- move_row_to_colnames(metro_raw)
  
  # 6. Clean + simplify column names
  metro_clean <- metro_tidy %>%
    select(FULL1_NAME, FULL2_NAME, MOVEDIN, MOVEDOUT) %>%
    rename(
      metro_current = FULL1_NAME,
      metro_1y      = FULL2_NAME,
      moved_in      = MOVEDIN, 
      moved_out     = MOVEDOUT
    ) %>%
    mutate(
      # Extract principal state from metro names
      state_current = pull_state_from_metro(metro_current),
      state_1y      = pull_state_from_metro(metro_1y),
      # Clean up
      across(c(moved_in, moved_out), as.integer)
    ) %>%
    filter(
      !is.na(metro_current), 
      !is.na(metro_1y),
      metro_current != metro_1y  # Exclude stayers
    )
  
  return(metro_clean)
}

# TEST FUNCTION
metro_flows <- metro_to_metro_migration()

# VERIFICATION
cat("Rows:", nrow(metro_flows), "Columns:", ncol(metro_flows), "\n")
cat("Sample states:", unique(metro_flows$state_current)[1:5], "\n")
#head(metro_flows, 10)
```

```{r task5-verify, message=FALSE}
library(gt)

# Load metro data
metro_flows %>%
  slice_head(n = 10) %>%
  gt() %>%
  cols_label(
    metro_current = "Current Metro",
    metro_1y = "Previous Metro", 
    moved_in = "Moved In",
    moved_out = "Moved Out",
    state_current = "Current State",
    state_1y = "Previous State"
  ) %>%
  fmt_integer(columns = c(moved_in, moved_out)) %>%
  tab_header(
    title = "Metro-to-Metro Migration Flows (2016-2020 ACS)",
    subtitle = "Sample of first 10 flows from Census API"
  ) %>%
  cols_width(
    metro_current ~ px(200),
    metro_1y ~ px(200)
  ) %>%
  cols_align(
    align = "left",
    columns = c(metro_current, metro_1y)
  ) %>%
  cols_align(
    align = "right", 
    columns = c(moved_in, moved_out, state_current, state_1y)
  ) %>%
  tab_options(
    table.font.size = px(14),
    data_row.padding = px(8),
    table.width = px(900)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

::: 



### Exploratory Data Analysis {.tabset .tabset-pills .unnumbered}


### States with the highest net population growth {.unnumbered}
**Q1.Which states have had the highest net population growth rates over the past decade?**
```{r}
#| label: tbl-state-growth-final
#| tbl-cap: "Top 10 States by Net Population Growth Rate (2015-2024)" 
#| echo: false
#| warning: false

library(tidyverse)
library(gt)
library(scales)

# Use your existing state_pop_df (2015-2019, 2021-2024)
state_growth <- state_pop_df %>%
  # Calculate growth rates between consecutive years
  arrange(state_name, year) %>%
  group_by(state_name) %>%
  mutate(
    pop_prev = lag(total_population),
    growth_rate = (total_population - pop_prev) / pop_prev,
    # Compound annual growth rate over full period =cagr
    years_span = max(year) - min(year),
    cagr = ((last(total_population) / first(total_population)) ^ (1/years_span) - 1)
  ) %>%
  # Summarize by state (overall CAGR or average annual growth)
  summarise(
    first_pop = first(total_population),
    last_pop = last(total_population),
    total_growth = last_pop - first_pop,
    total_growth_pct = (last_pop - first_pop) / first_pop,
    avg_annual_growth = mean(growth_rate, na.rm = TRUE),
    cagr = unique(cagr),
    n_years = n(),
    .groups = "drop"
  ) %>%
  # Focus on 50 states only (exclude territories)
  filter(!str_detect(state_name, "Puerto Rico|Guam|Virgin Islands")) %>%
  # Rank by CAGR (most appropriate for "growth rates over decade")
  arrange(desc(cagr)) %>%
  slice_head(n = 10)

# Create nicely formatted table
state_growth %>%
  select(state_name, first_pop, last_pop, total_growth_pct, cagr, n_years) %>%
  mutate(
    first_pop = comma(first_pop, accuracy = 1),
    last_pop = comma(last_pop, accuracy = 1),
    total_growth_pct = percent(total_growth_pct, accuracy = 0.1),
    cagr = percent(cagr, accuracy = 0.1),
    n_years = as.character(n_years)
  ) %>%
  gt() %>%
  fmt_markdown(columns = state_name) %>%
  cols_label(
    state_name = "State",
    first_pop = "2015 Pop",
    last_pop = "2024 Pop",
    total_growth_pct = "Total Growth",
    cagr = "CAGR",
    n_years = "Years"
  ) %>%
  tab_header(
    title = "Highest Population Growth Rates (2015-2024)",
    subtitle = "Compound Annual Growth Rate (CAGR) from ACS 1-Year Estimates"
  ) %>%
  cols_width(
    state_name ~ px(120),
    everything() ~ px(110)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_options(
    table.font.size = px(14),
    data_row.padding = px(6)
  )
```
**A**
The states with the highest population growth rates over the past decade were Idaho, Utah, Florida, Texas, and Nevada, with Idaho leading at about 20.9% total growth and a 2.1% Compound annual growth rate (CAGR). Overall, the table shows that many of the fastest-growing states were in the Mountain West and the south.


### Migration into and Out of New York {.unnumbered}
**Q2.If you meet someone who moved to New York State in the last year, what are the most likely states they moved from? If one of your friends announces they are moving out of the state, where are they most likely to move?**
```{r}
#| label: fig-ny-sankey-better-format
#| fig-cap: "New York Migration Flows (2023-2024)"
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"

library(tidyverse)
library(networkD3)
library(htmlwidgets)
library(htmltools)

migration_data <- state_to_state_migration()

ny_in <- migration_data %>%
  filter(state_current == "NEW YORK", state_1y != "NEW YORK") %>%
  group_by(state_1y) %>%
  summarise(value = sum(population, na.rm = TRUE), .groups = "drop") %>%
  slice_max(value, n = 5, with_ties = FALSE) %>%
  arrange(value) %>%
  mutate(from = state_1y, to = "NEW YORK") %>%
  select(from, to, value)

nodes_in <- tibble(
  name = c(ny_in$from, "NEW YORK"),
  group = c(ny_in$from, "NEW YORK")
)

links_in <- ny_in %>%
  transmute(
    source = match(from, nodes_in$name) - 1,
    target = match(to, nodes_in$name) - 1,
    value = value
  )

ny_out <- migration_data %>%
  filter(state_1y == "NEW YORK", state_current != "NEW YORK") %>%
  group_by(state_current) %>%
  summarise(value = sum(population, na.rm = TRUE), .groups = "drop") %>%
  slice_max(value, n = 5, with_ties = FALSE) %>%
  arrange(value) %>%
  mutate(from = "NEW YORK", to = state_current) %>%
  select(from, to, value)

nodes_out <- tibble(
  name = c("NEW YORK", ny_out$to),
  group = c("NEW YORK", ny_out$to)
)

links_out <- ny_out %>%
  transmute(
    source = match(from, nodes_out$name) - 1,
    target = match(to, nodes_out$name) - 1,
    value = value
  )

all_states <- unique(c(nodes_in$name, nodes_out$name))

palette_vals <- c(
  "#111111",
  "#4E79A7", "#F28E2B", "#59A14F", "#E15759", "#76B7B2",
  "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
)

state_colors <- setNames(rep_len(palette_vals, length(all_states)), all_states)
state_colors["NEW YORK"] <- "#111111"

colour_scale <- paste0(
  'd3.scaleOrdinal()',
  '.domain([', paste(sprintf('"%s"', names(state_colors)), collapse = ","), '])',
  '.range([', paste(sprintf('"%s"', unname(state_colors)), collapse = ","), '])'
)

style_sankey <- '
function(el, x) {
  d3.select(el).selectAll(".link")
    .style("stroke-opacity", 0.45)
    .style("stroke", "#9a9a9a");

  d3.select(el).selectAll(".node rect")
    .style("stroke", "#4f4f4f")
    .style("stroke-width", "0.6px");

  d3.select(el).selectAll(".node text")
    .style("font-size", "11.5px")
    .style("font-family", "Arial, sans-serif")
    .style("fill", "black");
}
'

p_in <- sankeyNetwork(
  Links = as.data.frame(links_in),
  Nodes = as.data.frame(nodes_in),
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  NodeGroup = "group",
  colourScale = colour_scale,
  fontSize = 12,
  nodeWidth = 36,
  nodePadding = 24,
  width = 900,
  height = 460,
  margin = list(top = 15, right = 140, bottom = 15, left = 20),
  sinksRight = TRUE
) %>% onRender(style_sankey) %>%
  prependContent(
    tags$div(
      style = "text-align:center; margin-bottom:6px; font-weight:700; font-size:16px; color:black;",
      "Into New York"
    )
  )

p_out <- sankeyNetwork(
  Links = as.data.frame(links_out),
  Nodes = as.data.frame(nodes_out),
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  NodeGroup = "group",
  colourScale = colour_scale,
  fontSize = 12,
  nodeWidth = 36,
  nodePadding = 28,
  width = 900,
  height = 460,
  margin = list(top = 15, right = 140, bottom = 15, left = 20),
  sinksRight = TRUE
) %>% onRender(style_sankey) %>%
  prependContent(
    tags$div(
      style = "text-align:center; margin-bottom:6px; font-weight:700; font-size:16px; color:black;",
      "Out of New York"
    )
  )

browsable(
  tagList(
    tags$div(style = "width:100%; margin-bottom:28px;", p_in),
    tags$div(style = "width:100%;", p_out)
  )
)
```
**A** If I met someone who moved to New York State in the last year, they would have likely moved from Delaware. Alabama, Colorado, Arkansas, and Alaska. Conversely, if one of my friends were to move out of the NY state, they would likely move to West Virginia, New Jersey, Vermont, Wisconsin, or Colorado. 


### Migration into and out of New York City {.unnumbered}
**Q3.If you meet someone who moved to New York City in the last year, what are the most likely metro areas they moved from? If one of your friends announces they are moving out of the city, where are they most likely to move?**

```{r}
#| label: fig-nyc-metro-sankey
#| fig-cap: "New York Metro Migration Flows (2016-2020 ACS). Note: Excludes international migration flows to focus on U.S. metro-to-metro patterns."
#| echo: false
#| warning: false
#| message: false
#| out-width: "100%"

library(tidyverse)
library(networkD3)
library(htmlwidgets)
library(htmltools)

data <- metro_flows

ny_metro_name <- data$metro_current[
  str_detect(data$metro_current, regex("New York|NYC", ignore_case = TRUE))
][11]

short_metro <- function(x) {
  x %>%
    str_replace(",?\\s+Metro Area$", "") %>%
    str_replace("New York-Newark-Jersey City, NY-NJ-PA", "New York") %>%
    str_replace("Los Angeles-Long Beach-Anaheim, CA", "Los Angeles") %>%
    str_replace("Philadelphia-Camden-Wilmington, PA-NJ-DE-MD", "Philadelphia") %>%
    str_replace("Washington-Arlington-Alexandria, DC-VA-MD-WV", "Washington") %>%
    str_replace("Boston-Cambridge-Newton, MA-NH", "Boston") %>%
    str_replace("Poughkeepsie-Newburgh-Middletown, NY", "Poughkeepsie") %>%
    str_replace("Bridgeport-Stamford-Norwalk, CT", "Bridgeport") %>%
    str_replace("Miami-Fort Lauderdale-Pompano Beach, FL", "Miami") %>%
    str_replace("Outside Metro Area within U.S. or Puerto Rico", "Outside metro")
}

ny_in <- data %>%
  filter(
    metro_current == ny_metro_name,
    metro_1y != ny_metro_name,
    metro_1y != "Outside Metro Area within U.S. or Puerto Rico",
    !str_detect(
      metro_1y,
      regex("Africa|Asia|Europe|Caribbean|Central America|South America|Northern America", ignore_case = TRUE)
    ),
    !is.na(moved_in)
  ) %>%
  group_by(metro_1y) %>%
  summarise(value = sum(moved_in, na.rm = TRUE), .groups = "drop") %>%
  slice_max(value, n = 5, with_ties = FALSE) %>%
  arrange(value) %>%
  mutate(from = short_metro(metro_1y), to = "New York metro") %>%
  select(from, to, value)

nodes_in <- tibble(
  name = unique(c(ny_in$from, ny_in$to)),
  group = unique(c(ny_in$from, ny_in$to))
)

links_in <- ny_in %>%
  transmute(
    source = match(from, nodes_in$name) - 1,
    target = match(to, nodes_in$name) - 1,
    value = value
  )

ny_out <- data %>%
  filter(
    metro_1y == ny_metro_name,
    metro_current != ny_metro_name,
    metro_current != "Outside Metro Area within U.S. or Puerto Rico",
    !str_detect(
      metro_current,
      regex("Africa|Asia|Europe|Caribbean|Central America|South America|Northern America", ignore_case = TRUE)
    ),
    !is.na(moved_out)
  ) %>%
  group_by(metro_current) %>%
  summarise(value = sum(moved_out, na.rm = TRUE), .groups = "drop") %>%
  slice_max(value, n = 5, with_ties = FALSE) %>%
  arrange(value) %>%
  mutate(from = "New York metro", to = short_metro(metro_current)) %>%
  select(from, to, value)

nodes_out <- tibble(
  name = unique(c(ny_out$from, ny_out$to)),
  group = unique(c(ny_out$from, ny_out$to))
)

links_out <- ny_out %>%
  transmute(
    source = match(from, nodes_out$name) - 1,
    target = match(to, nodes_out$name) - 1,
    value = value
  )

all_metros <- unique(c(nodes_in$name, nodes_out$name))

palette_vals <- c(
  "#666666",
  "#4E79A7", "#F28E2B", "#59A14F", "#E15759", "#76B7B2",
  "#EDC948", "#B07AA1", "#FF9DA7", "#9C755F", "#BAB0AC"
)

metro_colors <- setNames(rep_len(palette_vals, length(all_metros)), all_metros)
metro_colors["New York metro"] <- "#666666"

colour_scale <- paste0(
  'd3.scaleOrdinal()',
  '.domain([', paste(sprintf('"%s"', names(metro_colors)), collapse = ","), '])',
  '.range([', paste(sprintf('"%s"', unname(metro_colors)), collapse = ","), '])'
)

style_sankey <- '
function(el, x) {
  d3.select(el).selectAll(".link")
    .style("stroke-opacity", 0.45)
    .style("stroke", "#9a9a9a");

  d3.select(el).selectAll(".node rect")
    .style("stroke", "#4f4f4f")
    .style("stroke-width", "0.6px");

  d3.select(el).selectAll(".node text")
    .style("font-size", "12px")
    .style("font-family", "Arial, sans-serif")
    .style("fill", "black");
}
'

p_in <- sankeyNetwork(
  Links = as.data.frame(links_in),
  Nodes = as.data.frame(nodes_in),
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  NodeGroup = "group",
  colourScale = colour_scale,
  fontSize = 12,
  nodeWidth = 40,
  nodePadding = 28,
  width = 900,
  height = 500,
  margin = list(top = 10, right = 180, bottom = 10, left = 20),
  sinksRight = TRUE
) %>% onRender(style_sankey) %>%
  prependContent(
    tags$div(
      style = "text-align:center; margin-bottom:8px; font-weight:700; font-size:18px; color:black;",
      "Into New York metro"
    )
  )

p_out <- sankeyNetwork(
  Links = as.data.frame(links_out),
  Nodes = as.data.frame(nodes_out),
  Source = "source",
  Target = "target",
  Value = "value",
  NodeID = "name",
  NodeGroup = "group",
  colourScale = colour_scale,
  fontSize = 12,
  nodeWidth = 40,
  nodePadding = 28,
  width = 900,
  height = 500,
  margin = list(top = 10, right = 180, bottom = 10, left = 20),
  sinksRight = TRUE
) %>% onRender(style_sankey) %>%
  prependContent(
    tags$div(
      style = "text-align:center; margin-bottom:8px; font-weight:700; font-size:18px; color:black;",
      "Out of New York metro"
    )
  )

browsable(
  tagList(
    tags$div(style = "width:100%; margin-bottom:30px;", p_in),
    tags$div(style = "width:100%;", p_out)
  )
)
```

**A**If you meet someone who moved to New York City in the last year, they most likely came from the Philadelphia-Camden-Wilmington metro, followed by Boston-Cambridge-Newton, Washington-Arlington-Alexandria, Los Angeles-Long Beach-Anaheim, and Poughkeepsie-Newburgh-Middletown.

If someone is moving out of New York City, they are most likely heading to Philadelphia-Camden-Wilmington, then Boston-Cambridge-Newton, Washington-Arlington-Alexandria, Los Angeles-Long Beach-Anaheim, or Poughkeepsie-Newburgh-Middletown.

### State migration {.unnumbered}
**Q4.Which states have had the highest amounts of in-migration and out-migration? Furthermore, which states have had the highest amount of net migration?**

Absolute numbers, not per capita. Texas/Florida would lead if showing all 50 states.
```{r}
#| label: fig-state-migration-summary
#| fig-cap: "State Migration Summary (2023 to 2024, 2-Year Total)"
#| fig-width: 18
#| fig-height: 7
#| echo: false
#| warning: false
#| fig-align: center

library(tidyverse)
library(scales)
library(patchwork)

label_millions <- label_number(scale = 1e-6, suffix = "M", accuracy = 0.1)

# Your combined 2-year data (already deduplicated)
migration_clean <- state_to_state_migration() %>%
  filter(state_current != state_1y, !is.na(population)) %>%
  group_by(state_1y, state_current) %>%
  summarise(total_movers = sum(population), .groups = "drop")

cat("Total movers (2023+2024):", format(sum(migration_clean$total_movers), big.mark = ","), "\n")

# 1. TOP INFLOW STATES (2023+2024)
top_inflow <- migration_clean %>%
  group_by(state_current) %>%
  summarise(inflow = sum(total_movers), .groups = "drop") %>%
  slice_max(inflow, n = 5, with_ties = FALSE) %>%
  mutate(
    state_current = str_to_title(state_current),
    state_current = fct_reorder(state_current, inflow),
    inflow_label = label_millions(inflow)
  )

p1 <- ggplot(top_inflow, aes(y = state_current, x = inflow)) +
  geom_col(fill = "#2d8a47", alpha = 0.85, width = 0.78) +
  geom_text(aes(label = inflow_label), hjust = -0.1, size = 3.8, fontface = "bold") +
  scale_x_continuous(labels = label_millions, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 5 Inflow States", x = "People Moving In") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

# 2. TOP OUTFLOW STATES
top_outflow <- migration_clean %>%
  group_by(state_1y) %>%
  summarise(outflow = sum(total_movers), .groups = "drop") %>%
  slice_max(outflow, n = 5, with_ties = FALSE) %>%
  mutate(
    state_1y = str_to_title(state_1y),
    state_1y = fct_reorder(state_1y, outflow),
    outflow_label = label_millions(outflow)
  )

p2 <- ggplot(top_outflow, aes(y = state_1y, x = outflow)) +
  geom_col(fill = "#c0392b", alpha = 0.85, width = 0.78) +
  geom_text(aes(label = outflow_label), hjust = -0.1, size = 3.8, fontface = "bold") +
  scale_x_continuous(labels = label_millions, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 5 Outflow States", x = "People Moving Out") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

# 3. TOP NET MIGRATION
net_migration <- migration_clean %>%
  group_by(state_current) %>%
  summarise(inflow = sum(total_movers), .groups = "drop") %>%
  full_join(
    migration_clean %>%
      group_by(state_1y) %>%
      summarise(outflow = sum(total_movers), .groups = "drop") %>%
      rename(state_current = state_1y),
    by = "state_current"
  ) %>%
  mutate(
    inflow = replace_na(inflow, 0),
    outflow = replace_na(outflow, 0),
    net = inflow - outflow,
    direction = if_else(net > 0, "Net Gain", "Net Loss"),
    state_current = str_to_title(state_current)
  ) %>%
  slice_max(abs(net), n = 5, with_ties = FALSE) %>%
  arrange(net) %>%
  mutate(
    state_current = fct_reorder(state_current, net),
    net_label = label_millions(net),
    hjust_val = if_else(net > 0, -0.1, 1.1)
  )

p3 <- ggplot(net_migration, aes(y = state_current, x = net, fill = direction)) +
  geom_col(width = 0.78, alpha = 0.85) +
  geom_vline(xintercept = 0, color = "grey40", linewidth = 1) +
  geom_text(aes(label = net_label, hjust = hjust_val), size = 3.8, fontface = "bold") +
  scale_x_continuous(labels = label_millions, expand = expansion(mult = c(0.08, 0.12))) +
  scale_fill_manual(values = c("Net Loss" = "#c0392b", "Net Gain" = "#2d8a47")) +
  labs(title = "Top 5 Net Migration", x = "Net Migration") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 11),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

p1 + p2 + p3 + plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

**A**The states with the highest in-migration were Colorado, Vermont, Hawaii, North Dakota, and South Carolina. The highest out-migration was in Alaska, Arkansas, Colorado, Alabama, and Delaware, while the biggest net migration gains were in Vermont, Colorado, and Hawaii.


### Migration patters at the metro level {.unnumbered}
**Q5.Which metro areas have had the highest amounts of total in-migration and out-migration? Furthermore, which metro areas have had the highest amount of net migration?**
```{r}
#| label: fig-metro-migration-summary
#| fig-cap: "Metro Area Migration Summary (2023 to 2024, 2-Year Total)"
#| fig-width: 18
#| fig-height: 6
#| echo: false
#| warning: false
#| fig-align: center

library(tidyverse)
library(scales)
library(patchwork)

label_thousands <- label_number(scale = 1e-3, suffix = "K", accuracy = 1)

# Clean metro data
metro_clean <- metro_flows %>%
  filter(metro_current != metro_1y, !is.na(moved_in), !is.na(moved_out)) %>%
  group_by(metro_1y, metro_current) %>%
  summarise(
    moved_in = sum(moved_in, na.rm = TRUE),
    moved_out = sum(moved_out, na.rm = TRUE),
    .groups = "drop"
  )

# 1. TOP 5 INFLOW
top_inflow <- metro_clean %>%
  group_by(metro_current) %>%
  summarise(inflow = sum(moved_in), .groups = "drop") %>%
  arrange(desc(inflow)) %>%
  slice_head(n = 5) %>%
  mutate(
    metro_label = str_wrap(metro_current, 25),
    metro_label = factor(metro_label, levels = rev(metro_label)),
    inflow_label = label_thousands(inflow)
  )

p1 <- ggplot(top_inflow, aes(y = metro_label, x = inflow)) +
  geom_col(fill = "#2d8a47", alpha = 0.85, width = 0.78) +
  geom_text(aes(label = inflow_label), hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_thousands,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Top 5 Inflow Metros",
    x = "People Moving In",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  )

# 2. BOTTOM 5 OUTFLOW
bottom_outflow <- metro_clean %>%
  group_by(metro_1y) %>%
  summarise(outflow = sum(moved_out), .groups = "drop") %>%
  arrange(outflow) %>%
  slice_head(n = 5) %>%
  mutate(
    metro_label = str_wrap(metro_1y, 25),
    metro_label = factor(metro_label, levels = rev(metro_label)),
    outflow_label = label_thousands(outflow)
  )

p2 <- ggplot(bottom_outflow, aes(y = metro_label, x = outflow)) +
  geom_col(fill = "#c0392b", alpha = 0.85, width = 0.78) +
  geom_text(aes(label = outflow_label), hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_thousands,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Bottom 5 Outflow Metros",
    x = "People Moving Out",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  )

# 3. TOP 5 LARGEST NET MIGRATION
net_migration <- metro_clean %>%
  group_by(metro_current) %>%
  summarise(inflow = sum(moved_in), .groups = "drop") %>%
  full_join(
    metro_clean %>%
      group_by(metro_1y) %>%
      summarise(outflow = sum(moved_out), .groups = "drop") %>%
      rename(metro_current = metro_1y),
    by = "metro_current"
  ) %>%
  mutate(
    inflow = replace_na(inflow, 0),
    outflow = replace_na(outflow, 0),
    net = inflow - outflow
  ) %>%
  arrange(desc(net)) %>%
  slice_head(n = 5) %>%
  mutate(
    metro_label = str_wrap(metro_current, 25),
    metro_label = factor(metro_label, levels = rev(metro_label)),
    net_label = label_thousands(net)
  )

p3 <- ggplot(net_migration, aes(y = metro_label, x = net)) +
  geom_col(fill = "#1f78b4", alpha = 0.85, width = 0.78) +
  geom_text(aes(label = net_label), hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_thousands,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Top 5 Net Migration Metros",
    x = "Net Migration",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
  )

p1 + p2 + p3
```

**A**The metro areas with the highest in-migration were Los Angeles-Long Beach-Anaheim, Dallas-Fort Worth-Arlington, New York-Newark-Jersey City, Washington-Arlington-Alexandria, and Phoenix-Mesa-Chandler. The metro areas with the highest net migration were Dallas-Fort Worth-Arlington, Woodlands-Sugar Land, Minneapolis-St. Paul-Bloomington, Atlanta-Sandy Springs-Alpharetta, and Phoenix-Mesa-Chandler.


### States with lowest in-migration {.unnumbered}
**Q6.Which states have the highest fraction of residents who lived there last year? (That is, which states have the lowest proportional total in-migration?)**
```{r}
#| label: fig-q6-state-retention
#| fig-cap: "Top 10 States: Highest Share of Residents Living in Same State 1 Year Ago (2023-2024)"
#| fig-width: 12
#| fig-height: 8
#| echo: false
#| warning: false

library(tidyverse)
library(scales)

# Q6: Filter FIRST, then analyze
q6_analysis <- migration_combined %>%
  # FILTER FIRST - before any grouping
  filter(state_current != "UNITED STATES") %>%
  # Now aggregate duplicate state pairs
  group_by(state_current, state_1y) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  # Filter again after grouping (catches any artifacts)
  filter(state_current != "UNITED STATES", state_current != "UNITED STATES5") %>%
  # Calculate retention metrics
  group_by(state_current) %>%
  summarise(
    stayed_same_state   = sum(population[state_current == state_1y], na.rm = TRUE),
    total_in_migrants   = sum(population[state_current != state_1y], na.rm = TRUE),
    total_residents     = sum(population, na.rm = TRUE),
    same_state_share    = stayed_same_state / total_residents,
    .groups = "drop"
  ) %>%
  arrange(desc(same_state_share))

# Top 10 table
q6_table <- q6_analysis %>%
  slice_head(n = 10) %>%
  mutate(
    same_state_share = percent(same_state_share, accuracy = 0.1),
    stayed_same_state = comma(stayed_same_state, accuracy = 1),
    total_in_migrants = comma(total_in_migrants, accuracy = 1)
  ) %>%
  select(
    state_current,
    stayed_same_state,
    total_in_migrants,
    same_state_share
  )

#q6_table

# Clean plot data (no totals)
q6_plot_data <- q6_analysis %>%
  slice_head(n = 10) %>%
  mutate(
    state_current = str_to_title(state_current),
    state_current = fct_reorder(state_current, same_state_share)
  )

ggplot(q6_plot_data, aes(x = same_state_share, y = state_current)) +
  geom_col(fill = "#2d8a47", alpha = 0.85, width = 0.7) +
  geom_text(
    aes(label = percent(same_state_share, accuracy = 0.1)),
    hjust = -0.1, size = 3.5, fontface = "bold"
  ) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "States with Lowest Proportional In-Migration",
    subtitle = "Highest % of current residents who lived in same state 1 year ago (2023-2024)",
    x = "Share of Current Residents in Same State Last Year",
    y = NULL,
    caption = "Source: U.S. Census State-to-State Migration Flows"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 1),
    panel.grid.major.x = element_blank(),
    axis.ticks.y = element_blank()
  )
```

**A** The states with the highest share of residents who lived there last year were Alabama, Puerto Rico, Alaska, Washington, and Oklahoma. In other words, these states had the lowest proportional in-migration, with Alabama at about 99.0% and Puerto Rico and Alaska both at 98.5%.


### Migration’s Share of the Growth {.unnumbered}
**Q7.Which state has the highest fraction of its population growth attributable to net internal migration? That is, determine the total population growth of each state and see what fraction of it corresponds to net migration (as opposed to deaths, births, or international migration).**
```{r}
#| label: fig-q7-migration-growth
#| fig-cap: "States Where Net Migration Drives Largest Share of Population Growth (2023-2024)"
#| fig-width: 12
#| fig-height: 8
#| echo: false
#| warning: false

library(tidyverse)
library(scales)

# Q7: CORRECTED - Net migration contribution to population
q7_analysis <- migration_combined %>%
  # Filter totals FIRST
  filter(state_current != "UNITED STATES") %>%
  # Aggregate duplicate state pairs (KEEP state_current)
  group_by(state_current, state_1y) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  filter(state_current != "UNITED STATES", state_current != "UNITED STATES5") %>%
  
  # Calculate by destination state (KEEP state_current)
  group_by(state_current) %>%
  summarise(
    total_current_residents = sum(population, na.rm = TRUE),
    total_in_migrants = sum(population[state_1y != state_current], na.rm = TRUE),
    stayers = sum(population[state_1y == state_current], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # Calculate net migration growth share
  mutate(
    net_migration_growth = total_in_migrants - stayers,
    migration_growth_share = net_migration_growth / total_current_residents
  ) %>%
  filter(net_migration_growth > 0) %>%  # Only growing states
  arrange(desc(migration_growth_share))

# Top 10 table
q7_table <- q7_analysis %>%
  slice_head(n = 10) %>%
  mutate(
    net_migration_growth = comma(net_migration_growth, accuracy = 1),
    total_current_residents = comma(total_current_residents, accuracy = 1),
    migration_growth_share = percent(migration_growth_share, accuracy = 0.1)
  ) %>%
  select(
    state_current,
    `Net Migration Growth` = net_migration_growth,
    `Total Residents` = total_current_residents,
    `Growth Share` = migration_growth_share
  )

#q7_table

# Plot
q7_plot_data <- q7_analysis %>%
  slice_head(n = 10) %>%
  mutate(
    state_current = str_to_title(state_current),
    state_current = fct_reorder(state_current, migration_growth_share)
  )

ggplot(q7_plot_data, aes(x = migration_growth_share, y = state_current)) +
  geom_col(fill = "#2d8a47", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = percent(migration_growth_share, accuracy = 0.1)), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_continuous(labels = percent_format(accuracy = 1), 
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Highest Fraction of Population Growth from Net Migration",
    subtitle = "(2023-2024: In-migrants beyond stayers / Total current residents)",
    x = "% of Current Population from Net Internal Migration",
    y = NULL,
    caption = "Source: U.S. Census State-to-State Migration Flows"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 1),
    panel.grid.major.x = element_blank(),
    axis.ticks.y = element_blank()
  )
```

**A** The state with the highest fraction of population growth coming from net internal migration was Vermont, at about 95.0%. After Vermont, the highest shares were Hawaii, Idaho, Colorado, and Iowa.

### Destination States and Migration Patterns {.unnumbered}
**Q8.Which state had the highest amount of migration into your state? Of the people who moved out of your state, what was the most popular destination state?**
```{r}
#| label: q8-nyc-metro-destinations
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(gt)

data <- metro_flows

ny_metro_name <- data$metro_current[
  str_detect(data$metro_current, regex("New York", ignore_case = TRUE))
][1]

# Most likely metro areas people moved TO New York from
ny_in <- data %>%
  filter(
    metro_current == ny_metro_name,
    metro_1y != ny_metro_name,
    !is.na(moved_in)
  ) %>%
  group_by(metro_1y) %>%
  summarise(people = sum(moved_in, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(people)) %>%
  slice_head(n = 5) %>%
  mutate(direction = "Moved into New York from")

# Most likely metro areas people moved FROM New York to
ny_out <- data %>%
  filter(
    metro_1y == ny_metro_name,
    metro_current != ny_metro_name,
    !is.na(moved_out)
  ) %>%
  group_by(metro_current) %>%
  summarise(people = sum(moved_out, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(people)) %>%
  slice_head(n = 5) %>%
  mutate(direction = "Moved out of New York to")

q8_table <- bind_rows(
  ny_in %>% transmute(Direction = direction, State = metro_1y, People = people),
  ny_out %>% transmute(Direction = direction, State = metro_current, People = people)
)

q8_table %>%
  gt() %>%
  fmt_integer(columns = People) %>%
  tab_header(title = "Most Common Migration Destinations for New York Metro")

```
**A**
After excluding international migration, the Philadelphia-Camden-Wilmington metro area was the top U.S. metro source of movers into and out of New York City.

### The State’s Largest Metro Area {.unnumbered}
**Q9.What is the largest metropolitan area in your state? (You can answer this just using background knowledge; you do not have to use Census data here.)**
```{r}
#| label: q9-largest-metro
#| echo: false

cat("The largest metropolitan area in New York State is the New York City metropolitan area, also called the New York-Newark-Jersey City metro area.")
```


### The Metro Network {.unnumbered}
**Q10.Are there any metro areas that have a particularly high connection to your state’s largest metro? (E.g., NYCers were over 15% of people moving into the Miami Metro area.)**

```{r}
#| label: q10-metro-connection
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(gt)

data <- metro_flows

ny_metro_name <- data$metro_current[
  str_detect(data$metro_current, regex("New York|NYC", ignore_case = TRUE))
][1]

# Top metro areas sending people into New York metro
q10_in <- data %>%
  filter(
    metro_current == ny_metro_name,
    metro_1y != ny_metro_name,
    !is.na(moved_in)
  ) %>%
  group_by(metro_1y) %>%
  summarise(moved_in = sum(moved_in, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(moved_in)) %>%
  slice_head(n = 5)

# Share of movers into selected metros coming from New York metro
q10_share <- data %>%
  filter(
    metro_1y != ny_metro_name,
    !is.na(moved_in)
  ) %>%
  group_by(metro_current) %>%
  summarise(total_in = sum(moved_in, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    data %>%
      filter(metro_1y == ny_metro_name, !is.na(moved_in)) %>%
      group_by(metro_current) %>%
      summarise(from_ny = sum(moved_in, na.rm = TRUE), .groups = "drop"),
    by = "metro_current"
  ) %>%
  mutate(
    from_ny = replace_na(from_ny, 0),
    ny_share = from_ny / total_in
  ) %>%
  filter(from_ny > 0) %>%
  arrange(desc(ny_share)) %>%
  slice_head(n = 5)

q10_share %>%
  mutate(ny_share = percent(ny_share, accuracy = 0.1)) %>%
  select(metro_current, from_ny, total_in, ny_share) %>%
  gt() %>%
  fmt_integer(columns = c(from_ny, total_in)) %>%
  cols_label(
    metro_current = "Metro Area",
    from_ny = "Movers from NYC Metro",
    total_in = "Total In-Migration",
    ny_share = "Share from NYC Metro"
  ) %>%
  tab_header(title = "Metro Areas Most Connected to New York City")

```

**A**
Among metro areas with the strongest ties to New York City, Trenton-Princeton, NJ had the highest share of movers from the NYC metro, followed by Poughkeepsie-Newburgh-Middletown, Bridgeport-Stamford-Norwalk, East Stroudsburg, and Kingston. This suggests that New York City’s strongest migration connections are mostly with nearby metro areas in the Northeast.



## Final Insights and Deliverable

```{r}
library(tidyverse)
library(scales)

valid_states <- c(state.name, "District of Columbia", "Puerto Rico") %>%
  toupper()

migration_2024 <- migration_combined %>%
  filter(year_current == 2024) %>%
  mutate(
    state_current = toupper(str_trim(state_current)),
    state_1y = toupper(str_trim(state_1y))
  ) %>%
  filter(
    state_current %in% valid_states,
    state_1y %in% valid_states,
    !is.na(population),
    population > 0
  )

pop_2023_t1 <- migration_2024 %>%
  group_by(state_1y) %>%
  summarise(pop_t1 = sum(population, na.rm = TRUE), .groups = "drop")

pop_2024_t2 <- migration_2024 %>%
  group_by(state_current) %>%
  summarise(pop_t2 = sum(population, na.rm = TRUE), .groups = "drop")

national_t1 <- sum(pop_2023_t1$pop_t1, na.rm = TRUE)
national_t2 <- sum(pop_2024_t2$pop_t2, na.rm = TRUE)
r_2024 <- (national_t2 - national_t1) / national_t1

cat("National 2023 (t1):", comma(national_t1), "\n")
cat("National 2024 (t2):", comma(national_t2), "\n")
cat("r_2024 =", percent(r_2024, accuracy = 0.1), "\n")

flows_2024 <- migration_2024 %>%
  group_by(state_current, state_1y) %>%
  summarise(flow = sum(population, na.rm = TRUE), .groups = "drop")

M_2024 <- flows_2024 %>%
  left_join(pop_2023_t1, by = "state_1y") %>%
  mutate(M_ij_2024 = flow / pop_t1) %>%
  select(state_i = state_current, state_j = state_1y, M_ij_2024)

pop_start <- pop_2024_t2 %>%
  rename(state = state_current, population = pop_t2) %>%
  arrange(state)

cat("M_2024 pairs:", nrow(M_2024), "\n")
cat("Starting states:", nrow(pop_start), "\n")
```


```{r}
states <- sort(unique(pop_start$state))

M_mat <- M_2024 %>%
  tidyr::complete(state_i = states, state_j = states, fill = list(M_ij_2024 = 0)) %>%
  arrange(state_i, state_j) %>%
  pivot_wider(names_from = state_j, values_from = M_ij_2024) %>%
  arrange(state_i)

M_mat_num <- M_mat %>%
  select(-state_i) %>%
  as.matrix()

current_pop <- pop_start$population
names(current_pop) <- pop_start$state

projections <- tibble(
  year = 2024,
  state = states,
  population = current_pop[states]
)

cat("Starting projection loop...\n")

for (yr in 2025:2030) {
  next_pop <- as.numeric(M_mat_num %*% current_pop)
  names(next_pop) <- states
  
  projections <- bind_rows(
    projections,
    tibble(year = yr, state = states, population = next_pop)
  )
  
  current_pop <- next_pop
  cat("Year", yr, "- Total pop:", comma(sum(next_pop, na.rm = TRUE)), "\n")
}

pop_2030_professor <- projections %>%
  filter(year == 2030) %>%
  mutate(population = round(population)) %>%
  arrange(desc(population))

cat("\n✅ MODEL COMPLETE:\n")
cat("-", nrow(projections), "projected populations\n")
cat("-", nrow(M_2024), "M_ij parameters used\n")
```

```{r}
N_DISTRICTS <- 435
MAX_SEATS <- 100

pop_2030_clean <- pop_2030_professor %>%
  filter(state %in% valid_states, population > 0)

initial_states <- pop_2030_clean %>%
  mutate(n_districts = 1L)

remaining_seats <- N_DISTRICTS - nrow(initial_states)

priority_values <- pop_2030_clean %>%
  expand_grid(seats = 1:MAX_SEATS) %>%
  mutate(priority = population / sqrt(seats * (seats + 1)))

extra_seats <- priority_values %>%
  slice_max(priority, n = remaining_seats, with_ties = FALSE) %>%
  count(state, name = "extra_seats")

final_apportionment <- initial_states %>%
  left_join(extra_seats, by = "state") %>%
  mutate(
    extra_seats = replace_na(extra_seats, 0L),
    n_districts = n_districts + extra_seats,
    pop_per_district = round(population / n_districts)
  ) %>%
  select(state, population, n_districts, pop_per_district) %>%
  arrange(desc(n_districts), desc(population))

ny_final <- final_apportionment %>% filter(state == "NEW YORK")

cat("\n🎯 FINAL GOVERNOR'S BRIEFING - PROFESSOR'S EXACT MODEL\n")
cat("════════════════════════════════════════════════════════════\n")
cat("NEW YORK STATE 2032 PROJECTION\n")
cat("Population (2030):", comma(ny_final$population), "\n")
cat("Congressional Seats:", ny_final$n_districts, "\n")
cat("Population per Seat:", comma(ny_final$pop_per_district), "\n")
cat("\nTop 5 States:\n")

final_apportionment %>%
  slice_head(n = 5) %>%
  mutate(
    population = comma(population),
    pop_per_district = comma(pop_per_district)
  ) %>%
  print(n = Inf)

```
To reach our goal of increasing internal migration and strengthening New York’s political influence, the campaign should focus on the metro areas that already have the strongest connection to the New York City metro. These include Trenton-Princeton, Poughkeepsie-Newburgh-Middletown, Bridgeport-Stamford-Norwalk, East Stroudsburg, and Kingston. Most of these places are in the Northeast and already have ties to New York through commuting, family networks, or past migration, which makes them promising targets. People in these metros are also more likely to already know New York well, so the campaign can build on that familiarity and focus on jobs, housing, and opportunity.

A good slogan for the campaign would be something practical and local, like “Come Back to New York” or “New York Still Has Room for You.” The point is to keep the message simple and focused on nearby metro areas with the strongest migration links, especially Philadelphia and the other Northeast metros that already send people to New York or receive people from New York. In other words, the campaign should target places where New York already has a relationship, because that gives the state the best chance of increasing in-migration in a realistic way.

Based on the metro areas we are targeting and the strength of the existing migration links, we would hope to generate a modest but meaningful increase in net in-migration, on the order of roughly 10Kto 25K additional residents over several years. That would not change New York overnight, but it could still matter at the margin for population growth and congressional apportionment.



```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 5
#| dpi: 96

library(gganimate)
library(gifski)

ny_proj <- projections %>%
  filter(state == "NEW YORK") %>%
  arrange(year) %>%
  mutate(
    year = as.numeric(year),
    pop_millions = population / 1e6
  )

p_ny_anim <- ggplot(ny_proj, aes(x = year, y = pop_millions, group = 1)) +
  geom_line(color = "#1f78b4", linewidth = 1.2) +
  geom_point(color = "#1f78b4", size = 3) +
  scale_x_continuous(breaks = 2024:2030, limits = c(2024, 2030)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1, suffix = "M")) +
  labs(
    title = "Projected New York Population",
    x = "Year",
    y = "Population in Millions"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) +
  transition_reveal(year) +
  ease_aes("linear")

animate(
  p_ny_anim,
  nframes = 80,
  fps = 10,
  width = 800,
  height = 500,
  units = "px",
  res = 96,
  renderer = gifski_renderer()
)
```


::: {.callout-note title="AI Usage Statement" collapse="true"}

Perplexity was used to help write the code in this project, but all non-code text was generated without the use of any Generative AI tools. Additionally, Perplexity was used to provide additional background information on the topic and to brainstorm ideas for the final open-ended prompt.

:::

------------------------------------------------------------------------

This work ©2025 by miggonzalez was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
